HO unification from object language to meta language

### 2024-04-01 paper/main.tex
update intro of maybe-beta

\noindent
The result of $A \Ue{} C$ is that the two \linketa share the same \lhs.
By unifying the two \rhs we get
$\sigma = \{ A~ \mapsto C, B \mapsto D ~\}$.
In turn, given the map $\mathbb{M}$, this second assignment is decompiled to
$\rho = \{~ X \mapsto Y ~\}$ as expected.

@@

TODO: say that maybe-eta also work in not(llambda)

@@

All the previous sections we have dealt with terms in \llambda, however, it is
often possible to work with terms in \maybebeta and wish to unify them. There
are situtation, for example, where the oracle has given  which $\beta$-reduction problems (\maybebeta) appears any
time we deal with a subterm $t = X t_1 \dots t_n$, where $X$ is flexible and the
list $[t_1 \dots t_n]$ in not in \llambda. This unification problem is not
solvable without loss of generality, since there is not a most general unifier.
If we take back the example given in \cref{sec:nutshell}, the unification $F a =
a$ admits two solutions for $F$: $\rho_1 = \{F \mapsto \lambda x.x\}$ and
$\rho_2 = \{F \mapsto \lambda \_.a\}$. Despite this, it is possible to work with
$\maybebeta$ if an oracle provides a substitution $\rho$ such that $\rho t$
falls again in the \llambda.

On the other hand, the \Ue is not designed to understand how the $\beta$-redexes
work in the object language. Therefore, even if we know that $F$ is assigned
to $\lambda x.x$, \Ue is not able to unify $F a$ with $a$. On the other hand,
the problem $F a = G$ is solvable by \Ue, but the final result is that $G$ is
assigned to $(\lambda x.x) a$ which breaks the invariant saying that the
substitution of the meta language does not generate terms outide \wellb{} (Property \ref{prop:nf}).

The solution to this problem is to modify the compiler such that any sub-term $t$
considered as a potential $\beta$-redex is replaced with a hole $h$ and a new dedicated
link, called \linkbeta.

\begin{elpicode}
  type link-beta tm -> tm -> link.
\end{elpicode}

\def\rhs{\ensuremath{rhs}\xspace}
\def\lhs{\ensuremath{lhs}\xspace}

This link carries two terms, the former representing the variable $h$ for the
new created hole and the latter containing the subterm $t$. As for the \linketa,
we will call $h$ and $t$ respectively the left hand side (\lhs)
and the right hand side (\rhs) of the \linkbeta.

@@

\paragraph{Detection of \maybebeta} TODO: \dots

\paragraph{Compilation with \linkbeta}

In order to build a \linkbeta, we need to adapt the compiler so that it can
recognize these ``problematic'' subterms. The following code snippet illustrate
such behavior, we suppose the rule to be added just after \cref{rule:comp-pf}.

>>>>

\noindent
The result of $A \Ue{} C$ is that the two \linketa share the same \lhs.
By unifying the two \rhs we get
$\sigma = \{ A~ \mapsto C, B \mapsto D ~\}$.
In turn, given the map \mapStore, this second assignment is decompiled to
$\rho = \{~ X \mapsto Y ~\}$ as expected.

@@

@@

Until now, we have only dealt we unification of terms in \llambda. However, we
want the unification relation to be more robust so that it can work with terms
in \maybebeta. Unification in \maybebeta is in general a non-decidable
procedure, e.g. $X \appsep a \Uo a$ is a unification problem in \maybebeta,
since we have $X a$ has the shape $X t_1\dots t_n$ where $X$ is a unification
variable and $t_1\dots t_n$ is not a list of distinct names (in our example, $a$
is a constant, hence not a name). We also point out that this unification
problem admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$.

It is the case, however, given a list of unification $n$ problems \foUnifPb, if
$\foUnifPb_i$ is in \maybebeta, it is possible that the resolution of the
problems $\bigwedge_{j=0}^{i-1}\foUnifPb_j$ gives a partial substitution
$\rho$, such that $\rho\foUnifPb_i$ falls again in \llambda.
%
\printAlll
  {{{X,\lambda x.Y},
    {(X\appsep a),a}}}
  {{{A,\lambda x.B},
    {(A\appsep a),a}}}
  {{{Y,B,0},
    {X,A,0}}}
  {{}}

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$, can be solved in \llambda. On the other hand, even though \Uo is
able to solve \foUnifPb, \Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
not unifable problem $(\lambda x. B) \appsep a \Ue a$.

In order to encompass this unification, we capture terms $t$ in \maybebeta and
replace them with fresh varaibles $v$. As per \maybebeta, the variables $v$ and
the terms $t$ are linked through a \linkbeta.

\linkbeta guarantee \cref{inv:linklhs} and the term on the \rhs has the
following property:

\begin{invariant}[\linkbeta \rhs]
  The \rhs of any \linkbeta has the shape $X t_1\dots t_n$ such that $X$ is a
  flexible variable and $t_1\dots t_n$ is not in \llambda.
\end{invariant}

\linkbeta are put in \linkStore and activated when \rhs falls in \llambda.

% This property can be relaxed, to accept approximations: a dedicated
% section is given in a future section
\begin{corollary}
  If the \lhs of a \linkbeta is instantiated to a rigid term and its \rhs
  counterpart is still in \maybebeta, the current unification problem is not in
  \llambda and the unification fails.
\end{corollary}

\begin{proof}[Proof sketch]
  Given $X \appsep t_1\dots t_n \Ue t$ where $t$ is a rigid term and
  $t_1\dots t_n$ is not in \llambda. By construction, $X \appsep t_1\dots t_n$
  is replaced with a variable $V$, and the \linkbeta \linkbetaM{}{V}{X
  \appsep t_1\dots t_n} is created. The unification instantiates $V$ to $t$,
  making the \lhs of the link a rigid term, while \rhs is still in \maybebeta.
  The original problem is in fact outside \llambda and unification fails.
\end{proof}

@@

Detection of \maybebeta is quite simple to implement in the compiler, it is 
sufficient to

Dire che maybebeta is simple to detect e che maybe eta funziona bene and su 
termini che non sono il llambda

\begin{corollary}
  Right hand side is $X_{scope} t_1 \dots t_n$ and $t_1$ is in $scope$ or
  $t_1$ is not a name
\end{corollary}
